---
title: "Analysis"
author: "Erwin Lares"
affiliation: "Research Cyberinfrastructure at DoIT"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
knitr:
  opts_chunk: 
    message: false
    warning: false
params:
  data_file: "data/data.csv"
---

## The long game

Every coding project starts somewhere. Too often, that somewhere is a bare script — functional, but fragile. Decisions go undocumented, inputs go unexplained, and the gap between "it works on my machine" and "someone else can reproduce this" quietly widens.

This document argues for a better starting point: a Quarto project. By grounding your work in a `.qmd` file from the beginning, you gain a single source of truth that is human-readable, self-documenting, and built for reproducibility. The prose and the code evolve together, keeping your intentions visible alongside your implementation. And when the time comes to scale up — running your analysis on a computing cluster, sharing it with collaborators, or returning to it months later — the foundation is already in place.

The general workflow this document demonstrates:

-   Start with a Quarto document as the project blueprint
-   Write human-readable analysis with embedded, executable code
-   Derive a machine-friendly `.R` script automatically via `knitr::purl()`
-   Support execution across interactive sessions, `quarto render`, and `Rscript` from a single source file
-   Lay the groundwork for containerization and scaled execution when the time comes

## Why a Quarto document and not an Rscript file?

Most research software projects start as code. A script gets written, it works, and then the hard work begins — trying to remember why certain decisions were made, what the inputs mean, and what the output actually represents. Documentation, if it exists at all, is an afterthought. Collaboration becomes difficult. Reproducibility becomes a promise that is hard to keep.

There is a better way to start.

Quarto and R Markdown let you weave together prose, code, and output into a single document. By writing your workflow as a document first, you shift your attention to making your process human readable before it is machine executable. This is not just a stylistic preference — it is a fundamentally different way of thinking about research software. You are forced to articulate what you are doing and why, not just how. The result is a living document that serves as specification, implementation, and documentation all at once.

This approach has compounding benefits. New collaborators can read the document and understand the project without needing to reverse-engineer the code. Reviewers can follow the logic of your analysis alongside the code that implements it. Your future self will thank you when you return to the project six months later.

Crucially, choosing to start with a Quarto document does not mean sacrificing the machine-friendly code you need to actually run your analysis at scale. At the end of this document, a single step automatically extracts all the code blocks and writes them to a standalone `.R` file. That file can be containerized, shipped to a computing cluster, and executed without any of the document overhead. You maintain a single source of truth — the `.qmd` — and the machine-readable code is a derived artifact of it, always up to date, never out of sync.

The question worth asking is not whether your project is important enough to deserve this treatment. It is whether you can afford to start any other way.

## Limitations and tradeoffs

This approach works best for analyses that are relatively linear, self-contained, and where reproducibility and communication are as important as the computation itself. As a project grows in complexity — more data sources, longer runtimes, modular functions spread across multiple files — the single-document model starts to show its seams. Rendering overhead becomes a real cost, the derived `.R` file may need manual curation, and collaborators unfamiliar with Quarto face a steeper onramp than a plain script would require.

None of these are reasons to abandon the approach, but they are reasons to be intentional about when and how you apply it. Think of the Quarto document as the right starting point, not necessarily the permanent home for every piece of your project. As complexity grows, functions can migrate to separate `.R` files and be sourced in, keeping the document as the narrative spine while the implementation lives in more maintainable, testable units. The goal is not purity — it is keeping your work understandable, reproducible, and easy to hand off.

## Special considerations

This workflow will assume the following:

-   Your workflow is contained inside an R project.
-   Your project is using the `renv` package, so an updated `lockfile` exists within your project.
-   If you need to install additional packages, do so with `renv::install()` rather than `install.packages()` and make sure to update your lock file with `renv::snapshot()`.

## Toy analysis

One of the key advantages of starting your analysis in a Quarto document is that prose, code, and supporting visuals all live in the same place. You are not context-switching between a script and a separate README — the explanation and the implementation evolve together.

This includes diagrams. Before writing a single line of analysis code, you can sketch the shape of your workflow using a Mermaid or Graphviz diagram, giving collaborators and readers an orientation before they encounter the details. See @fig-mermaid for a bird's-eye view of what this analysis consists of.

```{mermaid}
%%| label: fig-mermaid
%%| fig-cap: "Analysis workflow "
flowchart LR
  A[load the libraries <br>and data] --> B[Prepare <br>the data]
  B --> C[Perform Calculations]
  C --> D[Print tabular results]
  C --> E[Render visualizations]
  D -->F[Save tabular results as .csv<br>Save plot as .png]
  E -->F
```

The analysis begins immediately below.

## Load Packages and Data

One of the subtle challenges of writing analysis code that lives in a Quarto document is that the same code may be executed in very different contexts: interactively in RStudio while you are developing, via quarto render when producing the final document, or via Rscript when running headlessly on a remote machine. Each context has a different and natural way of receiving inputs. The code below handles all three gracefully with a single, self-contained logic block. In an interactive session, the data path is hardcoded — convenient for development. When called from the command line via Rscript, the path is passed as a positional argument, which is how a job scheduler would supply it. When rendering the document with Quarto, the path falls back to the params block defined in the YAML header. The result is that data_file is always resolved to the right value regardless of how the code is being run, with no changes required between contexts. This makes the .qmd — and the .R file derived from it — genuinely portable.

```{r}
library(tidyverse)
library(gt)

data_file <- if (interactive()) {
  # Interactive session in RStudio: hardcoded path for development
  "data/data.csv"
} else {
  args <- commandArgs(trailingOnly = TRUE)
  if (length(args) >= 1) {
    # Rscript context: path passed as command-line argument
    args[1]
  } else {
    # Quarto render context: fall back to params
    params$data_file
  }
}

data <- read_csv(data_file)
args <- commandArgs(trailingOnly = TRUE)

```

## Tabular results

Sometimes you need numerical results as the output of your analysis. R is great for that.

The data we are using to motivate this analysis is the [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/) data set.

We are interested in calculating the traditional measures of central tendency (min, max, mean, and standard deviation) for all three species of penguins present in the data set.

The code chunk below accomplishes that by grouping the data into subsets by species. The four measures of central tendency are then calculated using the Base R functions `min`, `max`, `mean`, and `sd`.

```{r}
#| label: tbl-mot
#| tbl-cap: "Measures of tendency for 3 penguin species"

moft_bm <- data |>
    group_by(species) |>
    summarize(min_fl = min(flipper_length_mm, na.rm = TRUE),
              max_fl = max(flipper_length_mm, na.rm = TRUE),
              mean_fl = mean(flipper_length_mm, na.rm = TRUE),
              sd_fl = sd(flipper_length_mm, na.rm = TRUE))

t <- moft_bm |> 
  gt() |> 
  fmt_number(decimals = 0) |> 
  tab_header(
    title = "Measures of Tendency",
    subtitle = "Flipper Length in mm"
  )

# Renders table in Quarto/interactive contexts, suppress in Rscript to avoid Rplots.pdf

if (length(args) == 0) t


```

@tbl-mot shows that Adelie penguins are the smallest, Chinstrap penguins are the medium-sized species, and Gentoo penguins are the largest.

## Visualization results

As the old adage goes, "a plot is worth a thousand tables." The code chunk below creates a set of boxplots of flipper length in millimeters for each species of penguin.

```{r}
#| label: fig-boxplot
#| fig-cap: "Flipper length in mm"

p <- data |> 
    ggplot() +
    aes(x = flipper_length_mm, color = species) +
    geom_boxplot() + 
    labs(title = "Flipper length in mm") +
    theme_bw()

# Print plot in Quarto/interactive contexts, suppress in Rscript to avoid Rplots.pdf

if (length(args) == 0) p
```

@fig-boxplot shows boxplots derived from the data. Boxplots summarize the distribution of a variable by showing the median, the interquartile range (the middle 50% of the data), and the overall spread. Outliers, if present, appear as individual points beyond the whiskers. This makes boxplots particularly useful for comparing distributions across groups at a glance.

## Saving results

Although notebooks are a great way to share analysis, sometimes we might just want the results of the code — in our case, the table of aggregate measures of central tendency and the plot.

The code chunk below accomplishes that using the `write_csv()` function and the `ggsave()` function.

```{r}

moft_bm |> write_csv("results-folder/moft-bm.csv")

ggsave("results-folder/fl-mm-plot.png", plot = p)
```

## Saving the code

To recap, we have a document that is mostly written in language intended to be read by humans: it has formatted text to make it easier to read, contains images and diagrams to better convey the points we are trying to get across, *and* it contains the code that performs the actual calculations. As data scales up, it may make sense to avoid the document-rendering overhead, since at that stage we are interested in running the code rather than producing the narrative document.

We can leverage `knitr`, the engine that powers R in Quarto, to extract just the R code from the Quarto document using the `purl()` function. The code chunk below creates an `.R` file from the `.qmd`, placing it in the same directory and giving it the same base name.

```{r}
#| eval: false

#    knitr::purl("analysis.qmd", documentation = 1)

# knitr::purl() was moved to purl.R and is called as a Quarto post-render script
# via _quarto.yml. This ensures analysis.R is automatically regenerated from
# analysis.qmd on every successful render, eliminating the risk of code drift
# between the two files.
```

## Creating the ancillary files for command line execution 

As an analysis grows, it is often useful to split the input data into subsets — one per group — so that each subset can be processed independently. This is particularly valuable when preparing work for scaled execution, where individual files can be dispatched as separate jobs.

The code chunk below splits the data by species, writes each subset to its own `.csv` file, and saves a manifest of those filenames to `data/subdatasets.csv` for downstream use.

```{r}
#| eval: false

# grouping variable; would be a function argument in a package context
grouping_variable <- data$species

# split data into a named list by group
subsets <- split(data, grouping_variable)

# write one subset to disk as <name>.csv
save_sublist <- function(df, name){
  write_csv(df, paste0("data/", name, ".csv"))
}

# write all subsets to disk
walk2(subsets, names(subsets), save_sublist)

# save a filename manifest for downstream use
file_list <- tibble(filename = glue::glue("{names(subsets)}.csv"))
write_csv(file_list, "data/subdatasets.csv", col_names = FALSE)
```
