---
title: "From the Notebook to the Cluster"
subtitle: "Part 1: Start with the Document"
author: "Erwin Lares @ RCI"
affiliation: "Research Cyberinfrastructure at DoIT"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
knitr:
  opts_chunk: 
    message: false
    warning: false
params:
  data_file: "data/data.csv"
---

## The long game

This document walks through a set of structural decisions that transfer to any analysis you write: how to organize a project, how to resolve inputs across execution contexts, and how to derive a machine-readable script from a human-readable document. The Palmer Penguins dataset is just a vehicle. The patterns are the point — and when the time comes to scale up, they will have already done half the work for you.

## Why a Quarto document and not an Rscript file?

Most research projects start as a script. It works — until you need to share it, revisit it six months later, or explain why a decision was made. Documentation becomes an afterthought, and reproducibility becomes a promise that is hard to keep.

Quarto lets you write prose, code, and output in a single document. By writing the workflow as a document first, you articulate *what* you are doing and *why* alongside the code that implements it. New collaborators can follow the logic without reverse-engineering the script. Reviewers see the reasoning and the results together.

Crucially, this does not mean sacrificing a machine-executable artifact. At the end of this document, `knitr::purl()` extracts all R code chunks and writes them to a standalone `.R` file automatically on every render. That file can be containerized and run on a cluster without any document overhead. The `.qmd` is the source of truth; `analysis.R` is a derived artifact — always current, never manually maintained.

## Limitations and tradeoffs

This approach works best for analyses that are relatively linear, self-contained, and where reproducibility and communication are as important as the computation itself. As a project grows in complexity — more data sources, longer runtimes, modular functions spread across multiple files — the single-document model starts to show its seams. Rendering overhead becomes a real cost, the derived `.R` file may need manual curation, and collaborators unfamiliar with Quarto face a steeper onramp than a plain script would require.

None of these are reasons to abandon the approach, but they are reasons to be intentional about when and how you apply it. Think of the Quarto document as the right starting point, not necessarily the permanent home for every piece of your project. As complexity grows, functions can migrate to separate `.R` files and be sourced in, keeping the document as the narrative spine while the implementation lives in more maintainable, testable units. The goal is not purity — it is keeping your work understandable, reproducible, and easy to hand off.

## Special considerations

This workflow will assume the following:

-   Your workflow is contained inside an R project.
-   Your project is using the `renv` package, so an updated `lockfile` exists within your project.
-   If you need to install additional packages, do so with `renv::install()` rather than `install.packages()` and make sure to update your lock file with `renv::snapshot()`.

## Toy analysis

One of the key advantages of starting your analysis in a Quarto document is that prose, code, and supporting visuals all live in the same place. You are not context-switching between a script and a separate README — the explanation and the implementation evolve together.

This includes diagrams. Before writing a single line of analysis code, you can sketch the shape of your workflow using a Mermaid or Graphviz diagram, giving collaborators and readers an orientation before they encounter the details. See @fig-mermaid for a bird's-eye view of what this analysis consists of.

```{mermaid}
%%| label: fig-mermaid
%%| fig-cap: "Analysis workflow "
flowchart LR
  A[load the libraries <br>and data] --> B[Prepare <br>the data]
  B --> C[Perform Calculations]
  C --> D[Print tabular results]
  C --> E[Render visualizations]
  D -->F[Save tabular results as .csv<br>Save plot as .png]
  E -->F
```

The analysis begins immediately below.

## Load Packages and Data

One of the subtle challenges of writing analysis code that lives in a Quarto document is that the same code may be executed in very different contexts: interactively in RStudio while you are developing, via quarto render when producing the final document, or via Rscript when running headlessly on a remote machine. Each context has a different and natural way of receiving inputs. The code below handles all three gracefully with a single, self-contained logic block. 

In an interactive session, the data path is hardcoded — convenient for development. When called from the command line via Rscript, the path is passed as a positional argument, which is how a job scheduler would supply it. When rendering the document with Quarto, the path falls back to the params block defined in the YAML header. The result is that data_file is always resolved to the right value regardless of how the code is being run, with no changes required between contexts. This makes the .qmd — and the .R file derived from it — genuinely portable.

```{r}
library(tidyverse)
library(gt)

data_file <- if (interactive()) {
  # Interactive session in RStudio: hardcoded path for development
  "data/data.csv"
} else {
  args <- commandArgs(trailingOnly = TRUE)
  if (length(args) >= 1) {
    # Rscript context: path passed as command-line argument
    args[1]
  } else {
    # Quarto render context: fall back to params
    params$data_file
  }
}

data <- read_csv(data_file)

```

## Tabular results

Sometimes you need numerical results as the output of your analysis. R is great for that.

The data we are using to motivate this analysis is the [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/) data set.

We are interested in calculating a set of summary statistics — range (min and max), average (mean), and spread (standard deviation) — for all three species of penguins present in the data set.

The code chunk below accomplishes that by grouping the data into subsets by species. Each statistic is then calculated using the Base R functions `min`, `max`, `mean`, and `sd`.

```{r}
#| label: tbl-mot
#| tbl-cap: "Measures of tendency for 3 penguin species"

flipper_stats <- data |>
    group_by(species) |>
    summarize(min_fl = min(flipper_length_mm, na.rm = TRUE),
              max_fl = max(flipper_length_mm, na.rm = TRUE),
              mean_fl = mean(flipper_length_mm, na.rm = TRUE),
              sd_fl = sd(flipper_length_mm, na.rm = TRUE))

flipper_stats_tbl <- flipper_stats |> 
  gt() |> 
  fmt_number(decimals = 0) |> 
  tab_header(
    title = "Measures of Tendency",
    subtitle = "Flipper Length in mm"
  )

# Renders table in Quarto/interactive contexts, suppress in Rscript to avoid Rplots.pdf

if (length(args) == 0) flipper_stats_tbl

```

@tbl-mot shows that Adelie penguins are the smallest, Chinstrap penguins are the medium-sized species, and Gentoo penguins are the largest.

## Visualization results

As the old adage goes, "a plot is worth a thousand tables." The code chunk below creates a set of boxplots of flipper length in millimeters for each species of penguin.

```{r}
#| label: fig-boxplot
#| fig-cap: "Flipper length in mm"

p <- data |> 
    ggplot() +
    aes(x = flipper_length_mm, y = species, color = species) +
    geom_boxplot() + 
    labs(title = "Flipper length in mm",
         x = "Flipper length (mm)",
         y = NULL) +
    theme_bw()

# Print plot in Quarto/interactive contexts, suppress in Rscript to avoid Rplots.pdf

if (length(args) == 0) p
```

@fig-boxplot shows boxplots derived from the data. Boxplots summarize the distribution of a variable by showing the median, the interquartile range (the middle 50% of the data), and the overall spread. Outliers, if present, appear as individual points beyond the whiskers. This makes boxplots particularly useful for comparing distributions across groups at a glance.

## Saving results

The code chunk below writes the summary statistics table to a `.csv` and the plot to a `.png`.

```{r}

dir.create("results-folder", showWarnings = FALSE)

flipper_stats |> write_csv("results-folder/flipper-stats.csv")

ggsave("results-folder/fl-mm-plot.png", plot = p)
```


## Saving the code

When the analysis needs to run at scale, the document-rendering overhead is unnecessary — you want the code, not the narrative. `knitr::purl()` extracts all R code chunks from the `.qmd` and writes them to a standalone `.R` file with the same base name.

In this project, `purl()` is not called directly from within the document. Instead, it lives in a dedicated script — `purl.R` — that is registered as a post-render hook in `_quarto.yml`:

```yaml
project:
  post-render:
    - purl.R
```

This means that every time the document is rendered, `purl.R` runs automatically and regenerates `notebook2cluster-p1-start-with-the-document.R` from the current state of the `.qmd`. You never have to remember to extract the code manually, and the two files cannot drift out of sync as long as you render before you run.

The `documentation` argument controls how much of the prose carries over into the generated script. With `documentation = 1`, section comments are included as `#` comments, giving the `.R` file enough context to be readable on its own. With `documentation = 0`, only the bare code is written — useful if the script will be passed directly to a scheduler with no human reader in the loop.

```r
knitr::purl("notebook2cluster-p1-start-with-the-document.qmd", documentation = 1)
```

## Creating the ancillary files for command line execution 

As an analysis grows, it is often useful to split the input data into subsets — one per group — so that each subset can be processed independently. This is particularly valuable when preparing work for scaled execution, where individual files can be dispatched as separate jobs.

The code chunk below splits the data by species, writes each subset to its own `.csv` file, and optionally saves a manifest of those filenames to `data/manifest.csv` for downstream use.

```{r}

split_and_save <- function(data, grouping_var, write_manifest = TRUE) {
  
  # validate that grouping_var exists in data
  if (!grouping_var %in% names(data)) {
    stop(glue::glue("Column '{grouping_var}' not found in data."))
  }
  
  # split data into a named list by group
  subsets <- split(data, data[[grouping_var]])
  
  # write one subset to disk as <name>.csv
  save_subset <- function(df, name) {
    write_csv(df, paste0("data/", name, ".csv"))
  }
  
  # write all subsets to disk
  walk2(subsets, names(subsets), save_subset)
  
  # optionally write a manifest of created filenames
  if (write_manifest) {
    tibble(filename = glue::glue("{names(subsets)}.csv")) |>
      write_csv("data/manifest.csv", col_names = FALSE)
  }
  
  invisible(length(subsets))
}

split_and_save(data, grouping_var = "species", write_manifest = TRUE)
```

## Running the derived script from the command line

Once analysis.R has been generated and the data has been split into subsets, each subset can be processed independently by passing it as a command-line argument:

`Rscript notebook2cluster-p1-start-with-the-document.R data/Adelie.csv`

This is the same invocation that a job scheduler on an HTC or HPC cluster would use — one call per file, each running independently. At small scale you run it manually; at large scale you hand that same command to a submit script and let the scheduler parallelize across as many files as you have.

The fact that no changes are required between the two scenarios is the point. The portability was built in from the start — in the input-resolution logic in notebook2cluster-p1-start-with-the-document.qmd, carried forward automatically into notebook2cluster-p1-start-with-the-document.R. When the time comes to scale, the transition is a change in infrastructure, not a rewrite of the code.

## Final thoughts

The penguin data was never really the point. The summary statistics and boxplots were placeholders — concrete enough to demonstrate the pattern, simple enough to stay out of the way. What this document is actually about is the set of decisions made around the analysis: where to start, how to handle inputs across contexts, and how to keep the human-readable and machine-readable versions of your work from drifting apart. Those habits are worth building early, regardless of what the analysis is about.

